---
title: "Code and data from: 'Landfills as islands of grassland biodiversity: placing a next-generation habitat restoration plan in context.'"
author: "Mike Allen"
date: "7/27/2021"
output: html_document
---
# Background
This code compiles and maps data on closed and active landfills in the USA, Northeast / Mid-Atlantic region. It is based on the US EPA's [LMOP Landfill and Project Database](https://www.epa.gov/lmop/lmop-landfill-and-project-database) (USEPA 2021). It was used to create the following map and figure that appeared in Allen (Ecological Restoration, 2021): "Landfills as islands of grassland biodiversity: placing a next-generation habitat restoration plan in context."

# Load data and libraries 
```{r}
library(tidyverse) # version: tidyverse_1.3.1

# read in LMOP data from each state
nj <- read.csv("data/lmopdatanj.csv")
ny <- read.csv("data/lmopdatany.csv")
pa <- read.csv("data/lmopdatapa.csv")
ma <- read.csv("data/lmopdatama.csv")
ri <- read.csv("data/lmopdatari.csv")
ct <- read.csv("data/lmopdatact.csv")
de <- read.csv("data/lmopdatade.csv")
md <- read.csv("data/lmopdatamd.csv")

# combine all into one data frame
la_raw <- rbind(nj, ny, pa, ma, ri, ct, de, md); rm(nj, ny, pa, ma, ri, ct, de, md)

# clean variable names and select relevant variables
la <- la_raw %>% 
  mutate(tons_waste = 
           as.numeric(gsub(Waste.in.Place..tons., pattern = ",", replacement = ""))) %>%
  select(name = Landfill.Name, 
         lat = Latitude, 
         lon = Longitude, 
         State,
         County,
         City,
         tons_waste,
         open = Year.Landfill.Opened,
         closed = Landfill.Closure.Year,
         status =Current.Landfill.Status) %>%
  distinct() %>%
  # create closed/open/soon-to-close categories
  mutate(Status = case_when(status == "Closed" ~ "Closed",
                             status == "Open" & 
                              closed %in% 2021:2050 ~ 
                              "Closed by 2050",
                             TRUE ~ "Open"))

# make the dataframe into a spatial point object
la_sf <- la %>%
  mutate(latitude = lat, longitude = lon) %>%
  filter(is.na(latitude)==FALSE) %>%
  sf::st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# subset dataset into a smaller region
lr <- la %>%
  filter(State %in% c("NJ", "NY", "PA"))

# make that spatial
lr_sf <- la_sf %>%
  filter(State %in% c("NJ", "NY", "PA"))

# load boundaries of US states
states <- rnaturalearth::ne_states(returnclass = "sf", country = c("United States of America", "Canada"))
```

# Map of landfills
```{r}
ggplot(la_sf) +
  geom_sf(data = states) +
  geom_sf(aes(color = Status, shape = Status)) +
  geom_sf(data = filter(la_sf, grepl(name, pattern = "Croton")), 
          shape = "C", size = 10) +
  scale_color_manual(values = c("black", "gray", "white")) +
    coord_sf(xlim = c(-80.7, -69.5), ylim = c(37.5, 45.5), expand = FALSE) +
  theme_bw() +
  theme(legend.position = c(.8,0.2),
        legend.background = element_rect(fill = "darkgray", color = "black"),
        legend.key = element_rect(fill = "darkgray")) + 
  labs(shape = "Landfill status", color = "Landfill status")  +
  ggspatial::annotation_scale(
    location = "bl",
    bar_cols = c("grey60", "white")
  ) +
  ggspatial::annotation_north_arrow(
    location = "tr", which_north = "true",
    pad_x = unit(0.1, "in"), pad_y = unit(0.1, "in"),
    style = ggspatial::north_arrow_nautical(
      fill = c("grey40", "white"),
      line_col = "grey20"
    )
  )

# save plot
 ggsave("figures/landfill_locs.pdf", 
        height = 6, width = 6, dpi = 600)
 ggsave("figures/landfill_locs.png", 
        height = 6, width = 6, dpi = 300)

```
# Size of closed and soon-to-close landfills
```{r}

la %>%
  filter(status == "Closed" | closed %in% 2021:2050  ) %>%
ggplot() +
  geom_histogram(aes(x = tons_waste), color = "gray") +
  geom_vline(aes(xintercept = 10023234), color = "darkgray", linetype = 2, size = 2)  +
  annotate(geom = "text", x = 12023234, y = 30, 
           label = "Croton Point",
           color = "darkgray",
           hjust = 0) +
  scale_x_log10() +
  theme_bw() +
  theme(text = element_text(size = 13)) +
  labs(x = "Size of closed and soon-to-close landfills in the eastern US (tons)",
       y = "Number of landfills")

# save plot
ggsave("figures/landfill_size.pdf", 
       height = 6, width = 6, dpi = 600)
ggsave("figures/landfill_size.png", 
       height = 6, width = 6, dpi = 72)
```

# Session Info
R version 4.1.2 (2021-11-01)
Platform: x86_64-w64-mingw32/x64 (64-bit)
Running under: Windows 10 x64 (build 19043) 19043)

# References
Allen, M. C. (2021). Landfills as islands of grassland biodiversity: placing a next-generation habitat restoration plan in context. Ecological Restoration, 39:284-287.

US EPA (US Environmental Protection Agency). 2021. LMOP Landfill
and Project Database (accessed 28 July 2021). www.epa.gov/lmop/lmop-landfill-and-project-database.
